---
title: "Solar Power"
output: 
  html_document: 
    toc: yes
    keep_md: yes
    fig_caption: yes
    toc_depth: 5
    number_sections: yes
---

# intro

This is the May 3rd 2022 tidytuesday dataset exploration about solar power. It is my first tidytuesday since August 31 2021. I will start small by walking through some elementary steps using `R`. 

# Load libraries, get data 

This is how most data explorations begin

```{r setup, error=FALSE, warning=FALSE, message=FALSE}
#load libraries
library(tidytuesdayR)
library(tidyverse)
library(lubridate)

# get the data
capacity <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-03/capacity.csv')
wind <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-03/wind.csv')
solar <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-03/solar.csv')
average_cost <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-03/average_cost.csv')
```

## Explore the data

Let's start exploring!

###  average cost exploration 

First, have a look at the `average_cost` data itself. 
```{r}
average_cost
```
Now make a very simple line plot.

```{r}
average_cost %>%
  ggplot(aes(year,gas_mwh)) +
  geom_line()
```

I want to put all three quantities (gas, solar, and wind) on the same plot so that means reshaping the data to make a wide table into a long (or tall) table.

```{r}
average_cost_long<- average_cost %>%
  pivot_longer(
    cols = gas_mwh:wind_mwh, 
    names_to = "fuel_type", 
    values_to = "sourced_dollars_per_MWh",
    values_drop_na = TRUE) %>%
  mutate(fuel_type = str_replace_all(fuel_type,pattern="_mwh", replacement=""))

head(average_cost_long)
```
Now, a plot with all three fuel types.
```{r}
average_cost_long%>%
  ggplot(aes(year,sourced_dollars_per_MWh, color=fuel_type))+
  geom_line()
```

Plot cleanup. Give a title, label the axes, make the lines thicker.

```{r}
average_cost_long%>%
  ggplot(aes(year,sourced_dollars_per_MWh, color=fuel_type))+
  geom_line(size=2) +
  xlab("year")+
  ylab("Sourced $ per MWh")+
  ggtitle("Sourced $ per MWh for different fuel types")
```

Reduce clutter by labeling the y-axis with dollar signs, removing the x axis label, and renaming `fuel_type` to `fuel type` in the legend. 

```{r}
average_cost_long%>%
  ggplot(aes(year,sourced_dollars_per_MWh, color=fuel_type))+
  geom_line(size=2)+
  scale_y_continuous(labels=scales::dollar_format())+
  ggtitle("Sourced $ per MWh for different fuel types")+
  theme(axis.title.x=element_blank(), 
        #axis.text.x=element_blank(),
        #axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), 
        #axis.text.y=element_blank(),
        #axis.ticks.y=element_blank()
        )+
  labs(color="fuel type")
```


###  satisfactory plot 

Remove the gray background, add a marker for every year, and make the major y axis interval 25 with minor 5.

```{r}
average_cost_long%>%
  ggplot(aes(year,sourced_dollars_per_MWh, color=fuel_type))+
  geom_line(size=2)+
  scale_y_continuous(breaks=seq(0, 200, by=50),labels=scales::dollar_format())+
  scale_x_continuous(breaks=seq(2009, 2021, by=2))+
  ggtitle("Sourced $ per MWh for different fuel types")+
  theme(axis.title.x=element_blank(), 
        #axis.text.x=element_blank(),
        #axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), 
        #axis.text.y=element_blank(),
        #axis.ticks.y=element_blank()
        )+
  labs(color="fuel type")+
  theme_light()
```


## Explore the other data sets

###  capcity 

Starting with capacity, units are in gigawatts:

```{r}
capacity
```

Again, the data must be reshaped.
```{r}
capacity_long <- capacity %>%
  pivot_longer(
    cols=standalone_prior:hybrid_new,
    names_to = "category",
    values_to = "gigawatts",
    values_drop_na = TRUE)

capacity_long
```

Now a quick plot, which shows that `standalone_prior` has grown while `hybrid_new` and `hybrid_prior` have increased since 2017. 

```{r}
capacity_long %>%
  ggplot(aes(x=year,y=gigawatts,fill=category))+
  geom_bar(position = 'dodge',stat = 'identity')
```

Now let's do a side-by-side split multiple, split on new and prior.

```{r}
#separate(mytext, sep = " ", into = c("col1","col2"), remove = FALSE)

capacity_long_sep<- capacity_long %>%
  separate(category,sep="_",into = c("type","time"), remove = TRUE)

capacity_long_sep
```

```{r}
capacity_long_sep %>%
  ggplot(aes(x=year,y=gigawatts,fill=type))+
  geom_bar(position = 'dodge',stat = 'identity')+
  facet_wrap(~time)
```

###  solar

This is similar to the `wind` data.

```{r}
solar
```

A simple line graph of date and the estimated cost of megawatt per hour for solar.

```{r}
solar %>%
  ggplot(aes(date,solar_mwh)) +
  geom_line()
```

This data may lead itself better to dots instead.

```{r}
solar %>%
  ggplot(aes(date,solar_mwh)) +
  geom_point()
```

Visualizing solar capacity on the above graph as the size of the dot.

```{r}
solar %>%
  ggplot(aes(date,solar_mwh)) +
  geom_point(aes(size=solar_capacity))
```

Some color now.

```{r}
solar %>%
  ggplot(aes(date,solar_mwh)) +
  geom_point(aes(size=solar_capacity, color=solar_capacity))
```

###  satisfactory plot 

Too much overlap, so make the dots transparent.

```{r}
solar %>%
  ggplot(aes(date,solar_mwh)) +
  geom_point(aes(size=solar_capacity, color=solar_capacity, alpha=.5))+
  geom_smooth(method=loess)
```

###  taking it a bit too far

Change that moving average line to another color for fun.

```{r}
solar %>%
  ggplot(aes(date,solar_mwh)) +
  geom_point(aes(size=solar_capacity, color=solar_capacity, alpha=.5))+
  geom_smooth(method=loess,color="green", fill="red")
```

That's a but too much, but it looks cool. 

###  misc - will explore this later or discard?

Probably going to discard it?

```{r}
solar %>%
  mutate(year=year(solar$date))%>%
  ggplot(aes(solar_mwh,solar_capacity))+
  geom_point(aes(alpha=.5))
```

###  wind 

Take the nice dot plot with trendline (minus the colorful red trendline) and apply that to the wind dataset.

```{r}
wind %>%
  ggplot(aes(date,wind_mwh)) +
  geom_point(aes(size=wind_capacity, color=wind_capacity, alpha=.5))+
  geom_smooth(method=loess)
```

There must be a way to combine the `wind` and `solar` datasets.

```{r}
solar
```

```{r}
wind
```

###  Merge Wind and Solar Data

The dates don't match 1 to 1. Make them both long tables and then merge on dates. 

```{r}
wind_long <- wind %>%
  pivot_longer(
    cols = wind_mwh:wind_capacity,
    names_to = "category",
    values_to = "value",
    values_drop_na = TRUE
  )

wind_long
```

```{r}
solar_long <- solar %>%
  pivot_longer(
    cols = solar_mwh:solar_capacity,
    names_to = "category",
    values_to = "value",
    values_drop_na = TRUE
  )

solar_long
```

```{r}
wind_and_solar_long <- bind_rows(wind_long,solar_long)

wind_and_solar_long
```

Too much information on one plot, it's pretty but ultimately nonsense. 

```{r}
wind_and_solar_long %>%
  ggplot(aes(date,value)) +
  geom_point(aes(color=category, alpha=.5))+
  geom_smooth(method=loess)
```

Nope, we need to split up category.

```{r}
#separate(coltosplitgoeshere, sep = " ", into = c("col1","col2",...,"coln"), remove = TRUE)

wind_and_solar_long_sep <- wind_and_solar_long %>%
  separate(category,
           sep="_",
           into= c("power source","type"),
           remove=TRUE
           )

wind_and_solar_long_sep
```

###  conclusions 

It appears at the capacity increases over tie and costs per megawatt hour decrease with time. 

```{r}
wind_and_solar_long_sep %>%
  ggplot(aes(date,value))+
  geom_point(aes(color=`power source`))+
  facet_wrap(~type)
```

###  cost per megawatt hour for wind and solar

todo - add label and title, also add a plot like this for capcity

```{r}
wind_and_solar_long_sep %>%
  filter(type=="mwh") %>%
  ggplot(aes(date,value, color=`power source`))+
  geom_point(aes(alpha=.5))+
  geom_smooth()
```

###  satisfactory plot

```{r}
wind_and_solar_long_sep %>%
  filter(type=="mwh") %>%
  ggplot(aes(date,value, color=`power source`))+
  geom_point()+
  geom_smooth(method=`loess`)+
  scale_y_continuous(labels=scales::dollar_format())+
  ggtitle("Cost $ per MWh")+
  xlab("")+
  ylab("")+
  theme_light()+
  theme(legend.position="top")
```

###  aggregate 

Aggregate the above plot by year. 

```{r}
wind_and_solar_long_sep %>%
  filter(type=="mwh") %>%
  mutate(year = year(date)) %>%
  mutate(month = month(date)) %>%
  ggplot(aes(x=year, y = value))+
  geom_boxplot(
    aes(group = cut_width(year,1)),
        outlier.alpha=.5)+
  facet_grid(~`power source`)+
  scale_y_continuous(
    labels=scales::dollar_format()
    )+
  geom_smooth(alpha=.2)
```

